\Large Analyzing the Supreme Court Citation Network

### 2018-01-02 Tex/paper.tex
revising theory section

\textbf{Transitivity:} In a network of directed relations (e.g., A cites B, but B doesn't cite A) transitivity refers to the tendency for A to send a tie to C if A sends a tie to B and B sends a tie to C [CITE FROM SYLLABUS]. In undirected networks, transitivity is simply the process by which friends of friends become friends (i.e., a friend of a friend is a friend). The term, ``transitive closure'' refers to a tie forming from A to C in response to extant ties from A to B and B to C. When writing opinions, Supreme Court justices present the legal bases for their rulings, which often involves discussing the most primary/relevant precedents underpinning these legal bases, but also the precedents and legal rules on which the primary precedents were based. This process of presenting several layers/levels of precedent in an opinion follows the structure of transitive closure exactly---opinion A cites opinion B as a primary precedent, and then cites opinion C because opinion B cites opinion C. The two examples presented below illustrate this process.

The statute thus addresses the risk of a morally unjustifiable death sentence, not by minimizing it as precedent unmistakably requires, but by guaranteeing that in equipoise cases the risk will be realized, by ?placing a ?thumb [on] death?s side of the scale,? ? Sochor v. Florida, 504 U. S. 527, 532 (1992) (quoting Stringer v. Black, 503 U. S. 222, 232 (1992); alteration in original).

https://supreme.justia.com/cases/federal/us/548/163/dissent2.html

``Never before the decision in Union Gas had we suggested that the bounds of Article III could be expanded by Congress operating pursuant to any constitutional provision other than the Fourteenth Amendment. Indeed, it had seemed fundamental that Congress could not expand the jurisdiction of the federal courts beyond the bounds of Article III. Marbury v. Madison, 1 Cranch 137 (1803). The plurality's citation of prior decisions for support was based upon what we believe to be a misreading of precedent. See Union Gas, 491 U. S., at 40-41 (SCALIA, J., dissenting). The plurality claimed support for its decision from a case holding the unremarkable, and completely unrelated, proposition that the States may waive their sovereign immunity, see id., at 14-15 (citing Parden v. Terminal Railway of Ala. Docks Dept., 377 U. S. 184 (1964)), and cited as precedent propositions that had been merely assumed for the sake of argument in earlier cases, see 491 U. S., at 15 (citing Welch v. Texas Dept. of Highways and Public Transp., 483 U. S., at 475-476, and n. 5, and County of Oneida v. Oneida Indian Nation of N. Y., 470 U. S., at 252).''

https://supreme.justia.com/cases/federal/us/517/44/case.html

\textbf{Reciprocity:} Reciprocity (also referred to as mutuality) is the tendency for node B to send a tie to node A in response to or coordination with A sending a tie to B [CITE FROM SYLLABUS]. It is typically not possible for reciprocated ties to form in legal opinions. Most citations reference past opinions that were issued before the citing opinion's case was even argued before the Court.  However, opinions written within the same Supreme Court term are often drafted in tandem, and can cite each other reciprocally.  Opinion A citing opinion B within the same term represents a signal that opinion B is relevant to the legal reasoning underpinning opinion A. Unlike the citations themselves, the applicability of legal rules or lines of reasoning across cases is not directed---if A is relevant to B, B is highly likely to be relevant to A. We expect opinions written within the same term to exhibit a high degree of reciprocity. Below we provide two example passages from opinions that illustrate the phenomenon of within-term reciprocity.

On a more specific level, Western argues that flight engineers must meet the same stringent qualifications as pilots, and that it was therefore quite logical to extend to flight engineers the FAA's age 60 retirement rule for pilots. Although the FAA's rule for pilots, adopted for safety reasons, is relevant evidence in the airline's BFOQ defense, it is not to be accorded conclusive weight. Johnson v. Mayor and City Council of Baltimore, ante at 472 U. S. 370-371. The extent to which the rule is probative varies with the weight of the evidence supporting its safety rationale and "the congruity between the . . . occupations at issue." Ante at 472 U. S. 371. In this case, the evidence clearly established that the FAA, Western, and other airlines all recognized that the qualifications for a flight engineer were less rigorous than those required for a pilot. [Footnote 28]

https://supreme.justia.com/cases/federal/us/472/400/case.html

The city, supported by several amici, argues for affirmance nonetheless. It asserts first that the federal civil service statute is not just a federal retirement provision unrelated to the ADEA, but in fact establishes age as a BFOQ for federal firefighters based on factors that properly go into that determination under the ADEA, see Western Air Lines, Inc. v. Criswell, post p. 472 U. S. 400. Second, the city asserts, a congressional finding that age is a BFOQ for a certain occupation is dispositive of that determination with respect to nonfederal employees in that occupation.

\url{https://supreme.justia.com/cases/federal/us/472/353/case.html#370}

\textbf{Popularity} Popularity, also termed ``preferential attachment'' is the tendency for ties to be sent to nodes to which many ties have already been sent [CITE FROM SYLLABUS]. Citations to an opinion signal both the Court's awareness of the legal reasoning of the case and the Court's evaluation that the opinion is an authoritative precedent. The more citations, the stronger this signal. Landmark cases, or those that establish new legal rules, are particularly authoritative and accrue citations from most future opinions that follow the respective line of reasoning. The passage below, from XXXX, in which an authoritative opinion is referenced, and even discussed in terms of the number of other cases by which it was followed. We consider each of these contentions in turn. Based on this popularity phenomenon, we expect that new citations will be disproportionately directed at cases to which many past citations have accrued.

"The first case in which this Court struck down a statute under the Equal Protection Clause of the Fourteenth Amendment was Strauder v. West Virginia, 100 U. S. 303, decided in the 1879 Term. [Footnote 2/1] In the 1961 Term, we squarely held that the manner of apportionment of members of a state legislature raised a justiciable question under the Equal Protection Clause, Baker v. Carr, 369 U. S. 186. That case was followed by numerous others, e.g.: that one person could not be given twice or 10 times the voting power of another person in a state-wide election merely because he lived in a rural area..."

https://supreme.justia.com/cases/federal/us/400/112/case.html

\textbf{Sociality} Sociality is the tendency for ties sent to beget more ties sent [CITE]. Structurally, this means that there are few nodes that send a moderate number of ties---the number of ties sent is either small or fairly large. In terms of Supreme Court opinions,  for each opinion discussed, that discussion is likely to raise other tangential issues on which the justices will want to draw upon past opinions.  Furthermore, for each opinion that applies to and is cited by the current opinion, there is often a case/opinion that needs to be discussed in terms of why it does not apply. Justices often clarify not only those legal principles that apply, but often those that do not. The passages below represent examples of this process of citation proliferation, or sociality.

@@

	In its raw form a citation is a link between two documents, but it is the result of human judgment about the utility of the cite \citep{fowler2008authority}. Studies of citation networks can be broadly categorized according to whether they treat this set of observed behavioral patterns as information that can be used to understand different aspects of the system, or treat them as outcomes to be modeled. We discuss these in turn.

@@

% Network Statistics
The c-TERGM is specified through 
the decision regarding which network statistics $h(\cdot)$ to incorporate. We include the following statistics for the Supreme Court citation network, which are derived from the interdependence hypotheses described in the previous section:
$$h_{edges}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to \sum_{ij} C_{ij}, $$
the number of citations.  $h_{edges}$ performs the function of an intercept, and models the expected value of any given edge.
$$h_{outstar}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to \sum_{i=1}^N {\sum_{j \neq i} c_{ij} \choose 2} $$
the number of out-two-stars. An out-two-star is a configuration in which one node sends to two other nodes. The number of these configurations grows quadratically as the origins of ties concentrate on a few highly active/social senders. The out-two-stars configuration is commonly used to model sociality in ERGMs.
$$h_{instar}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C \to \sum_{j=1}^N {\sum_{i \neq j} c_{ij} \choose 2} $$
the number of in-two-stars. An in-two-star is a configuration in which one node receives ties from two other nodes. The number of these configurations grows quadratically as the destinations of ties concentrate on a few highly popular recipients. The out-two-stars configuration is commonly used to model popularity in ERGMs.
\begin{align*}
h_{triangle}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to & \sum_{j<i<k} C_{ij}\cdot C_{jk} \cdot C_{ki} + C_{ji}\cdot C_{jk} \cdot C_{ki} + C_{ij}\cdot C_{kj} \cdot C_{ki} \\ & + C_{ji}\cdot C_{kj} \cdot C_{ki} + C_{ij}\cdot C_{jk} \cdot C_{ik} + C_{ji}\cdot C_{jk} \cdot C_{ik}  \\ & + C_{ij}\cdot C_{kj} \cdot C_{ik} + C_{ji}\cdot C_{kj} \cdot C_{ik},  
\end{align*}
The number of triangles in the network.  Triangles measure triad closure, as discussed above. $$h_{covariate}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to \sum_{ij} C_{ij}X_{ij}, $$ the effect of an exogenous covariate ($X$). We include several exogenous covariates based on this standard statistic formulation. These covariates are discussed below.

@@

    (g) Sociality & (h) Popularity \\
\includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/o2star_coef_trend.pdf} & \includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/i2star_coef_trend.pdf} \\

@@

>>>>

\textbf{Transitivity:} In a network of directed relations (e.g., A cites B, but B doesn't cite A) transitivity refers to the tendency for A to send a tie to C if A sends a tie to B and B sends a tie to C \citep{holland1971transitivity}. In undirected networks, transitivity is simply the process by which friends of friends become friends (i.e., a friend of a friend is a friend). The term, ``transitive closure'' refers to a tie forming from A to C in response to extant ties from A to B and B to C. When writing opinions, Supreme Court justices present the legal bases for their rulings, which often involves discussing the most primary/relevant precedents underpinning these legal bases, but also the precedents and legal rules on which the primary precedents were based. This process of presenting several layers/levels of precedent in an opinion follows the structure of transitive closure exactly---opinion A cites opinion B as a primary precedent, and then cites opinion C because opinion B cites opinion C. The two examples presented below illustrate this process.

In the first example, a passage from Kansas v. Marsh (548 U.S. 163, 2006)---a case considering the constitutionality of a death sentence statute in Kansas. In this example, the case Stringer v. Black is cited by Kansas v. Marsh as a case that is quoted by Sochor v. Florida. The primary precedent under discussion in this passage of the opinion is Sochor v. Florida, but Stringer v. Black is cited as a result of its role in the Sochor v. Florida opinion.
\begin{quotation}
The statute thus addresses the risk of a morally unjustifiable death sentence, not by minimizing it as precedent unmistakably requires, but by guaranteeing that in equipoise cases the risk will be realized, by ``placing a `thumb [on] death's side of the scale,' '' Sochor v. Florida, 504 U. S. 527, 532 (1992) (quoting Stringer v. Black, 503 U. S. 222, 232 (1992); alteration in original).
\end{quotation} %https://supreme.justia.com/cases/federal/us/548/163/dissent2.html
The second example is a passage from Seminole Tribe of Fla. v. Florida, 
(517 U.S. 44 1996)---a case addressing the rights of groups and citizens to sue states in federal court. In this example, Pennsylvania v. Union Gas Co
(491 U.S. 1, 1989) is the primary precedent under discussion, and several cases are cited and discussed in terms of their roles as precedents in the Union Gas opinion. 
\begin{quotation}
Never before the decision in Union Gas had we suggested that the bounds of Article III could be expanded by Congress operating pursuant to any constitutional provision other than the Fourteenth Amendment. Indeed, it had seemed fundamental that Congress could not expand the jurisdiction of the federal courts beyond the bounds of Article III. Marbury v. Madison, 1 Cranch 137 (1803). The plurality's citation of prior decisions for support was based upon what we believe to be a misreading of precedent. See Union Gas, 491 U. S., at 40-41 (SCALIA, J., dissenting). The plurality claimed support for its decision from a case holding the unremarkable, and completely unrelated, proposition that the States may waive their sovereign immunity, see id., at 14-15 (citing Parden v. Terminal Railway of Ala. Docks Dept., 377 U. S. 184 (1964)), and cited as precedent propositions that had been merely assumed for the sake of argument in earlier cases, see 491 U. S., at 15 (citing Welch v. Texas Dept. of Highways and Public Transp., 483 U. S., at 475-476, and n. 5, and County of Oneida v. Oneida Indian Nation of N. Y., 470 U. S., at 252).'
\end{quotation}% https://supreme.justia.com/cases/federal/us/517/44/case.html

\noindent \textbf{Reciprocity:} Reciprocity (also referred to as mutuality) is the tendency for node B to send a tie to node A in response to or coordination with A sending a tie to B \citep{garlaschelli2004patterns}. It is typically not possible for reciprocated ties to form in legal opinions. Most citations reference past opinions that were issued before the citing opinion's case was even argued before the Court.  However, opinions written within the same Supreme Court term are often drafted in tandem, and can cite each other reciprocally.  Opinion A citing opinion B within the same term represents a signal that opinion B is relevant to the legal reasoning underpinning opinion A. Unlike the citations themselves, the applicability of legal rules or lines of reasoning across cases is not directed---if A is relevant to B, B is highly likely to be relevant to A. We expect opinions written within the same term to exhibit a high degree of reciprocity. Below we provide two example passages from opinions that illustrate the phenomenon of within-term reciprocity.

The first case in our example reciprocal dyad is a passage from Western Air Lines v. Criswell (472 U.S. 400, 1985)---a case considering mandatory retirement in the context of age discrimination laws. The second case in the dyad, Johnson v. Mayor of Baltimore (472 U.S. 353, 1985) is another case considering whether mandatory retirement violates the Age Discrimination in Employment Act. These cases addressed very similar legal questions, which increased the likelihood that they would inform each other, and the opinions were written within the same term, which made it possible for them to cite each other. 
\begin{quotation}
{\em From Western Air Lines:} On a more specific level, Western argues that flight engineers must meet the same stringent qualifications as pilots, and that it was therefore quite logical to extend to flight engineers the FAA's age 60 retirement rule for pilots. Although the FAA's rule for pilots, adopted for safety reasons, is relevant evidence in the airline's BFOQ defense, it is not to be accorded conclusive weight. Johnson v. Mayor and City Council of Baltimore, ante at 472 U. S. 370-371. The extent to which the rule is probative varies with the weight of the evidence supporting its safety rationale and "the congruity between the . . . occupations at issue." Ante at 472 U. S. 371. In this case, the evidence clearly established that the FAA, Western, and other airlines all recognized that the qualifications for a flight engineer were less rigorous than those required for a pilot.
\\~\\
{\em From Johnson:} The city, supported by several amici, argues for affirmance nonetheless. It asserts first that the federal civil service statute is not just a federal retirement provision unrelated to the ADEA, but in fact establishes age as a BFOQ for federal firefighters based on factors that properly go into that determination under the ADEA, see Western Air Lines, Inc. v. Criswell, post p. 472 U. S. 400. Second, the city asserts, a congressional finding that age is a BFOQ for a certain occupation is dispositive of that determination with respect to nonfederal employees in that occupation. 
\end{quotation} % https://supreme.justia.com/cases/federal/us/472/400/case.html \url{https://supreme.justia.com/cases/federal/us/472/353/case.html#370}

\noindent \textbf{Popularity} Popularity, also termed ``preferential attachment'' is the tendency for ties to be sent to nodes to which many ties have already been sent \citep{barabasi1999emergence}. Citations to an opinion signal both the Court's awareness of the legal reasoning of the case and the Court's evaluation that the opinion is an authoritative precedent. The more citations, the stronger this signal. Landmark cases, or those that establish new legal rules, are particularly authoritative and accrue citations from most future opinions that follow the respective line of reasoning. The passage below, from Oregon v. Mitchell (400 U.S. 112, 1970)---a case on the legality of state age restrictions on voting in federal elections---illustrates this popularity dynamic. In this opinion passage Baker v. Carr is cited in reference to its role as a landmark precedent, and noted for the number of other cases by which it has been followed.  which an authoritative opinion is referenced, and even discussed in terms of the number of other cases by which it was followed. 
\begin{quotation}
The first case in which this Court struck down a statute under the Equal Protection Clause of the Fourteenth Amendment was Strauder v. West Virginia, 100 U. S. 303, decided in the 1879 Term. [Footnote 2/1] In the 1961 Term, we squarely held that the manner of apportionment of members of a state legislature raised a justiciable question under the Equal Protection Clause, Baker v. Carr, 369 U. S. 186. That case was followed by numerous others, e.g.: that one person could not be given twice or 10 times the voting power of another person in a state-wide election merely because he lived in a rural area..."
\end{quotation} %https://supreme.justia.com/cases/federal/us/400/112/case.html

\textbf{Activity} Activity, or sender activation, is the tendency for ties sent to beget more ties sent \citep{howard2016understanding}. Structurally, this means that there are few nodes that send a moderate number of ties---the number of ties sent is either small or fairly large. In terms of Supreme Court opinions,  for each opinion discussed, that discussion is likely to raise other tangential issues on which the justices will want to draw upon past opinions.  Furthermore, for each opinion that applies to and is cited by the current opinion, there is often a case/opinion that needs to be discussed in terms of why it does not apply. Justices often clarify not only those legal principles that apply, but often those that do not. The passages below represent examples of this process of citation proliferation, or activation.

@@

[HERE]	In its raw form a citation is a link between two documents, but it is the result of human judgment about the utility of the cite \citep{fowler2008authority}. Studies of citation networks can be broadly categorized according to whether they treat this set of observed behavioral patterns as information that can be used to understand different aspects of the system, or treat them as outcomes to be modeled. We discuss these in turn.

@@

% Network Statistics
The c-TERGM is specified through 
the decision regarding which network statistics $h(\cdot)$ to incorporate. We include the following statistics for the Supreme Court citation network, which are derived from the interdependence hypotheses described in the previous section:
$$h_{edges}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to \sum_{ij} C_{ij}, $$
the number of citations.  $h_{edges}$ performs the function of an intercept, and models the expected value of any given edge.
$$h_{outstar}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to \sum_{i=1}^N {\sum_{j \neq i} c_{ij} \choose 2} $$
the number of out-two-stars. An out-two-star is a configuration in which one node sends to two other nodes. The number of these configurations grows quadratically as the origins of ties concentrate on a few highly active/social senders. The out-two-stars configuration is commonly used to model activity (sender activation) in ERGMs.
$$h_{instar}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C \to \sum_{j=1}^N {\sum_{i \neq j} c_{ij} \choose 2} $$
the number of in-two-stars. An in-two-star is a configuration in which one node receives ties from two other nodes. The number of these configurations grows quadratically as the destinations of ties concentrate on a few highly popular recipients. The out-two-stars configuration is commonly used to model popularity in ERGMs.
\begin{align*}
h_{triangle}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to & \sum_{j<i<k} C_{ij}\cdot C_{jk} \cdot C_{ki} + C_{ji}\cdot C_{jk} \cdot C_{ki} + C_{ij}\cdot C_{kj} \cdot C_{ki} \\ & + C_{ji}\cdot C_{kj} \cdot C_{ki} + C_{ij}\cdot C_{jk} \cdot C_{ik} + C_{ji}\cdot C_{jk} \cdot C_{ik}  \\ & + C_{ij}\cdot C_{kj} \cdot C_{ik} + C_{ji}\cdot C_{kj} \cdot C_{ik},  
\end{align*}
The number of triangles in the network.  Triangles measure triad closure, as discussed above. $$h_{covariate}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to \sum_{ij} C_{ij}X_{ij}, $$ the effect of an exogenous covariate ($X$). We include several exogenous covariates based on this standard statistic formulation. These covariates are discussed below.

@@

    (g) Activity & (h) Popularity \\
\includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/o2star_coef_trend.pdf} & \includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/i2star_coef_trend.pdf} \\

@@

\section{conclusion}

MAJOR LIMITATION---DO NOT CONSIDER SIGN OF CASES
